---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

```{r setoptions, echo=FALSE}
library(knitr)
opts_chunk$set(echo = TRUE)
```


## Loading and preprocessing the data
1. Set up directory for downloading data
```{r}
dir.create("./data", showWarnings = FALSE)

downloadURL <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
downloadedFile <- "./data/activity.zip"
```

2. Download and unzip data
```{r}
if(!file.exists(downloadedFile)) {
    download.file(downloadURL, downloadedFile, method = "curl")
    unzip(downloadedFile, exdir = "./data")
}
```

3. Load data to environment
```{r}
df <- read.csv(file = "./data/activity.csv", header = TRUE)
```

4. Convert date feature to DateTime format
```{r}
df$date <- as.Date(df$date, "%Y-%m-%d")
```

5. Take a quick look of our data
```{r}
head(df)
```


## What is mean total number of steps taken per day?
1. Calculate the total number of steps taken per day
```{r}
total.steps <- tapply(df$step, df$date, sum)
```

2. Histogram of the total number of steps taken per day
```{r}
library(ggplot2)
qplot(total.steps,
      geom = "histogram",
      bins = 10,
      main = "Total steps taken per day",
      xlab = "Steps")
```

3. The mean and median of the total number of steps taken per day

*Mean:*
```{r}
mean.steps <- tapply(df$steps, df$date, mean)
summary(mean.steps)
```
The minimum value for the mean of the total number steps is very low with 0.14. Otherwise, the maximum value is 73.59 which is very high value for this figure.

*Median:*
```{r}
median.steps <- tapply(df$steps, df$date, median)
summary(median.steps)
```

The median values for all of the dates are the same with 0. 
It means, for each date, there are more than 50% of the time that the enthusiasts did not take any steps.


## What is the average daily activity pattern?



## Imputing missing values



## Are there differences in activity patterns between weekdays and weekends?
